\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-02-09 doc/publications/plos16/plos16.tex
BMD section added notes

\cursedforest is included in our previously developed framework for Spark-based genomic data analysis,
VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
comprehensive analysis toolkit that can scale for future data demands, where the richness from WGS data can be
utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAS catalog and generate datasets
of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
compute paradigms to deal with these challenges.

In the first section we explore some of the properties of random forests as these will also be apparent in the
\cursedforest implementation. Here we provide demonstrations using synthetic data. Secondly, we demonstrate the ability of
\cursedforest to select previously demonstrated features from a bone mineral density (BMD) dataset. 
Thirdly, we demonstrate the scalability of \cursedforest in respect to the dimensionality of data by building a random
forest model on whole-genome data from the 1000 Genomes Project. Here we report the error in predicting ethnicity.
Finally, given the role different parameter values can play in model construction, we explore the effect that tuning these
parameters can have on the prediction accuracy of the model.

@@

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are invariant to monotonic scaling of the data;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to uninformative predictors;
  \item are able to capture interactions between features (which is important for modeling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

@@

The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labeled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
$$
I_G(f) = \sum_{q = 1}^Q f_q ( 1 - f_q ), 
$$
which is at a minimum when all observations at the node are in the same class.

@@

\subsubsection{Synthetic data} 
\label{section:synthetic_data}
 Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomic variation) randomly generated from a uniform distribution with equal
probabilities.

@@

This may be because a large \mtry\ generates more correlated trees as the same important features have a higher chance
of being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification
tasks where random features can mimic the response variable, hence resulting more in a performance plateau.  
\marginpar{unclear what this means}
Increasing
the number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelizing trees at
the node level hence caters perfectly to the requirement of large feature datasets as more trees can be built given a
fixed time budget.

@@

To test whether the associated features have biological meaning, we investigate whether variants in or near
BMD-associated genes are enriched are enriched toward the top of our list of our ranked variant list. To begin with,
we start with 26 known BMD genes~\cite{Duncan.2011}. We also include the 6 genes that Duncan {\it et al.} find with
a new BMD association. We then add to this list the genes that Estrada {\it et al.} found with a new BMD association.
In total, this gives us a list of 100 genes.

We label our contributing variants that fall within 100 kb of these 100 genes and observe that they are ranked
significantly higher than variants from other locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$),
supporting our hypothesis that biologically relevant features are contributing more information to our classification model.

@@

Investigating the lowly-ranked genes, we observe DCDC5 (Doublecortin Domain Containing 5) ranked 9,830 out of 10,000. This gene is 
one of the BMD-associated genes postulated by literature that Duncan {\it et al.}~\cite{Duncan.2011} inspect.
Similar to their findings, our results also indicate that this gene does not have a strong association with BMD. 
In fact, DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear \cite{Thakker2012}. 
Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.

@@

@@

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
large number of areas including variable selection in high dimensions. Consider
Fig~\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a
distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the
phase-transition line variable recovery is still possible by a combinatorial approach.

>>>>

\cursedforest is included in our previously developed framework for Spark-based genomic data analysis,
VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
comprehensive analysis toolkit that can scale for future data demands, where the richness from WGS data can be
utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAS catalogue and generate datasets
of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
compute paradigms to deal with these challenges.

In the first section we explore some of the properties of random forests as these will also be apparent in the
\cursedforest implementation. Here we provide demonstrations using synthetic data. Secondly, we demonstrate the ability of
\cursedforest to select previously demonstrated features from a bone mineral density (BMD) dataset. 
Thirdly, we demonstrate the scalability of \cursedforest in respect to the dimensionality of the data by building a random
forest model on whole-genome data from the 1000 Genomes Project. Here we report the error in predicting ethnicity.
Finally, given the role different parameter values can play in model construction, we explore the effect that tuning these
parameters can have on the prediction accuracy of the model.

@@

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are invariant to monotonic scaling of the data;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to uninformative predictors;
  \item are able to capture interactions between features (which is important for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

@@

The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labelled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
$$
I_G(f) = \sum_{q = 1}^Q f_q ( 1 - f_q ), 
$$
which is at a minimum when all observations at the node are in the same class.

@@

\subsubsection{Synthetic data} 
\label{section:synthetic_data}
 Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomics variation) randomly generated from a uniform distribution with equal
probabilities.

@@

This may be because a large \mtry\ generates more correlated trees as the same important features have a higher chance
of being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification
tasks where random features can mimic the response variable, hence resulting more in a performance plateau.  
\marginpar{unclear what this means}
Increasing
the number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelising trees at
the node level hence caters perfectly to the requirement of large feature datasets as more trees can be built given a
fixed time budget.

@@

To test whether the associated features have biological meaning, we investigate whether the known BMD genes investigated
by Duncan {\it et al.}~\cite{Duncan.2011} are enriched amongst the highly ranked genes. We therefore label all variants
that fall within 100 kb of these 32 BMD genes and observe that they are ranked significantly higher than other locations
in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant features are
contributing more information to our classification model.

@@

Investigating the lowly-ranked genes, we observe DCDC5 (Doublecortin Domain Containing 5) ranked 9,667 out of 10,000. This gene is 
one of the BMD-associated genes postulated by literature that Duncan {\it et al.}~\cite{Duncan.2011} inspect.
Similar to their findings, our results also indicate that this gene does not have a strong association with BMD. 
In fact, DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear \cite{Thakker2012}. 
Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.

@@

%1:693731 rs12238997
% rs1035665 A genome-wide association study of global gene expression http://www.nature.com/ng/journal/v39/n10/full/ng2109.html rs1035665 as master regulator of ATP Binding Cassette Subfamily F Member 2 ABCF1  with ABCF2 634 known 
% rs6702460 1	91536	91536	14	1.0296212827525244E-6
% rs2058553 2	20573002	20573002	15	9.896033178354981E-7

@@

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
large number of areas including variable selection in high dimensions. Consider
Fig~\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a
distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behaviour of a number of
regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the
phase-transition line variable recovery is still possible by a combinatorial approach.
